---
title: "R Boot Camp 1: Introduction to R"
author: "Jesse Brunner"
date: "`r format(Sys.Date())`"
output:
  pdf_document:
    toc: true
---

### What is R?  
We will be using R and R Studio in this class. In fact many of you have probably used R before in some context. But it still bares answering the questions, what is R?

R is an object-oriented scripting language that combines: 
*   a programming language called S, developed by John Chambers at Bell Labs, that can be used for numerical simulation of deterministic and stochastic dynamic models  
*   an extensive set of functions for classical and modern statistical data analysis and modeling  
*   graphics functions for visualizing data and model output, and creating publication-ready figures  
*   a user interface with a few basic menus and extensive help facilities.  

R is open source, cross platform, and free (in both contexts). R is very extensible (and has a _ton_ of packages extending it already). One good thing about this is that most anything you can think of has already been implemented in some package, somewhere. The down side is that there may be 12 ways to do a simple thing, leading to a lot of confusion. You can always "roll your own"---and we will---but it takes a fair bit of familiarity and understanding to do that. Unfortunately, the built-in help is made for more advanced R users and the bare-bones R is not particularly pretty, helpful, or coherent. Fortunately, we can address at least some of these issues by using R Studio.

### What is R Studio?
R Studio is an integrated development environment (IDE) that provides a coherent, pretty, and useful front-end to R. It allows you to see your console, script, output, files, etc., all in one place. It also has a lot of useful bells and whistles (code completion, version control, interactivity, save/publish as HTML or PDF) and this nice integration of Markdown, which is what I'm using to "knit" together text+code+results into this pretty document. We will be learning to use Markdown and R Studio along the way. It may feel like we're doing too much all at once, but there is really little downside and a lot to be gained by doing it "right" from the start.  


### First steps: R as a calculator
NOTE: For today, we are going to enter commands on the console, the window with a `>` in it. Type your commands after the `>` and then hit enter or return. 

```{r}
2+3
a <- 2+3 # save the output of the command to the variable (object), "a"
# the arrow means "assign"
a
```
Things to notice:

  1.  We can use basic math as in most any calculator, Excel, or stats program.  
  2.  We can assign the output of a command (here `2+3`) to a variable (here, named `a`).  
      *  All R assignments have this form: `objectName <- value`  (Read it as "a gets 2+5".)
      *  NOTE: You will see some others (e.g., Bolker's book) use an equal sign (`=`) to assign things to variables. While this works, it will cause problems down the line. So _please_ use the arrow (`<-`). Alt + - (the minus sign) is a shortcut you can use.  
  3.  You will see that everything to the right of the hash mark (technically, the "[octothorpe](http://99percentinvisible.org/episode/octothorpe/)") is ignored as a comment. We will be use a lot of comments to make it clear what we are doing and why.  

Anyway, we can do math on variables.
```{r}
a/3
b <- 3
a/b
a^b # all the basic math is there, and them some!
log(a) # natural log
log10(a) # log10
exp(a) # e^a
exp(a)/b
exp(a/b) # parentheses clarify order of operations
round( sqrt( exp(a/b) ) ) # We can do lots operations all at once.
round(         sqrt(exp(a/b    )
										)
							 ) # the "white space" is ignored. But keep it tidy!
# And R will give you a `+` prompt instead of the normal `>` when you hit return and 
# your parentheses or squiggly brackets don't add up.
```
Technically, `exp()`, `sqrt()`, `round()` and anything else with the name and parentheses are functions. So are `+`, `*`, `/`, and `^`, though they don't look like most other functions. So what we are doing is sending the results of one function (e.g., `exp(a/b)`) to another (e.g., `sqrt()`) to another (`round()`). The "arguments" of the function, things it takes and acts on, go inside the parentheses. For simple math we don't need to worry about this, but later on it will be helpful to know. We will spend more time on functions later, so don't sweat it.

Math and function can also work on **vectors**
```{r}
d <- c(-3, -2, 1, 5, 19, 7) # the function c() is for "concatenate" 
# or "string together"
d  # The variable "d" is thus a vector of values 
d/3 # math gets applied element-wise
d/b
d/pi # Note that "pi" is a protected name meaning 3.14159...
abs(d)
```
In many cases the function is applied element-by-element, but some functions return a single variable, e.g., 
```{r}
mean(d)  # functions can be applied to variables. This one takes the whole vector
summary(d)
min(d)
max(d)
```

### Beginning to work with data in R
In general we will import data into R (e.g., from a spreadsheet program), but it can be very useful to specify data or sequences in some instances. We saw that we could use the `c()` function to concatenate a string of numbers (or letters if we used quotation marks around each element). We can also generate sequences of integers using the `:` function (another one that doesn't look like a function),
```{r}
1:10 # == c(1,2,3,4,5,6,7,8,9,10)
10:1 # == c(10,9,8,7,6,5,4,3,2,1)
-5:10
e <- 1:20
e
```
or use the `seq()` function to be clearer about what we want.
```{r}
seq(from=1, to=10, length=10)
seq(from=1, to=10, by=1) # equivalent
seq(from=-5, to=10, length=10)
seq(from=-5, to=10, by=1) #not equivalent
f <- seq(from = -10, to = 20, length = 20)
f
```
-----------------

**An aside on functions** This use of the `seq()` function brings up an important issues about functions. Before when we were getting means or logs, we simply gave the function a number or vector of numbers (e.g., `a` or `d`). Here were are specifying the start (`from`) and end (`to`) and even how long (`length`) or by what steps (`by`) we want the sequence. Most functions are written to accept either the arguments in a particular order (here, from, to, by) or, as have just seen, by name. 

Try seeing what happens if you give the `seq()` function numbers without names and vice versa. Also try reversing the numbers.
```{r, eval=FALSE}
seq(1, 10)
seq(1,10, 0.5)
seq(1, 10, 20)
seq(20, 10, 1)
seq(from=1, to=10, length=20)
seq(length=20, from=1, to=10)
```
While we can (and often do) get away with just providing numbers (or names of variables) in the right order, it is much better practice to name the arguments. At a minimum this prevents you from making stupid mistakes, but it can also make your code much easier to read for me, a stranger, or your future self. 

-------------------

We can pick out particular elements of a vector by subsetting using vector notation
```{r}
f # this is the whole vector of numbers
f[1] # the first element
f[c(3,4,1)] # the third, fourth, and first elements of f
f[-c(3,4,1)] # everything but the third, fourth, and first elements
# Note the negative sign, which means "not" in this context
f[5:2] # the fifth through second elements of f
```

We can also use logical tests to pick out elements that meet certain requirements
```{r}
f > -1  # a logical test: are the elements of f greater than negative one? 
f[ f > -1 ] # Can use this logical test to pull out just the elements that are greater than -1
f <- f[ f > -1 ] # This will overwrite f with the elements that are greater than -1
f # see, it is now a smaller subset. 
# Note that the original version of f is gone. 
# The only way to get it back is to define it again
f <- seq(from=-10, to=20, length=20) # return it to it's original version
```

### Data frames
Most of the time you won't be working with lots of different vectors. In fact, most of the time this would be a very, very bad idea... You would have far too things to keep track of and keeping them all in sync would be a nightmare. Instead, we will almost always be working with data frames. These are basically the R version of a spreadsheet for nice, rectangular data. They:
  *  keep related variables all together  
  *  keep rows in sync so that when you filter by one variable, you filter out whole rows
  *  you can send (aka "pass") a data frame to _most_ functions with at `data = data_frame` argument.  
Also, when you import data it will almost always be imported from something like a spreadsheet and stored in R as a data frame. Let's see how data frames works as put these different elements together.


### Putting it all together: working with data
So now that we have the basics, let us play with some simple data to see how things all work together. First, let us enter the data. (You may want to just copy and paste this.)
```{r}
# light intensity in µE per m2 per second
Light <- c(20, 20, 20, 20, 21, 24, 44, 60, 90, 94, 101)  
# maximum growth rate of the green alga Chlorella vulgaris
rMax <- c(1.73, 1.65, 2.02, 1.89, 2.61, 1.36, 2.37, 2.08, 2.69, 2.32, 3.67) 

# put them into a data frame
growth <- data.frame(Light, rMax)

# Just to keep the vectors called Light and rMax from causing problems,
# Let's delete them
rm(Light, rMax)
```
These data represent the maximum growth rate of a green alga under various light intensities. How would you expect these to be related?

We can get some summary information and properties of our data frame. For such a small one this is overkill, but these are very handy with real, generally much larger data sets. 
```{r}
names(growth) # names of the columns
ncol(growth)
nrow(growth)
summary(growth) # summary of each variable (i.e., column)
str(growth) # this is short for "structure"
head(growth) # the first 6, by default, rows
tail(growth) # the last 6, by default, rows
```
You can select just one variable in a data frame with the `$` (i.e., subset a given row)
```{r}
growth$Light
growth$rMax
mean(growth$rMax)
```
And you can subset data frames with indices as we did before with vectors, but now we need to pay attention to the two dimensions, rows and columns. Note, a missing index means "all".
```{r}
growth[1, ] # first row, all columns
growth[ ,1] # first column, all rows
growth[3,2] # third row, second column
```
One last neat feature of data frames: you can add columns to them!
```{r}
growth$NewVar <- with(growth, 100*rMax/Light)
growth
```


Plotting the data is also recommended. We'll spend a lot more time graphing soon, so here is just a bare-bones example
```{r}
# plot rmax versus light
plot(x=growth$Light, y=growth$rMax)
```
Remember how I just said that most functions can take a data frame as an argument? Well `plot()` isn't one of them! This is frustrating, yes, but a) it helps illustrate how you can specify  variables in a data frame and b) we won't be using `plot` much more. 

Here is a nicer way to provide a data frame to a function that doesn't want to take it...
```{r}
with(data=growth, 
		 plot(x=Light, y=rMax))
```
The `with()` function basically calls the function in a new environment with the data. Or you can think about it as saying, "With these data, do this thing". Like, with this data frame, calculate a correlation coefficient between Light and rMax.
```{r}
with(data=growth,
		 cor.test(x=Light, y=rMax))
```

But again, most of the time the function can take the data frame. Consider a linear regression between rMax and Light. The `lm()` function (for linear model) and many other modeling functions use the formula notation of `y ~ x`, which can be read as y by x. Here we want to see how rMax changes with Light.
```{r}
# get the regression line
lm1 <- lm(rMax ~ Light, data=growth)
lm1
```
Notice that we do not get a whole lot out of the regression by itself, just the coefficients. To get the rest of the statitics we need to use the `summary()` function on the regression object, `lm1`. 
```{r}
summary(lm1)
```
There we go, the full regression statistics we're used to seeing. Spend a bit of time working your way through the output to be sure you see what you expect and know what things are. 

You can also get an ANOVA table from any linear model with the `anova()` function. 
```{r}
anova(lm1)
```
Note that `anova()` just computes the ANOVA table for a fitted model (i.e., the results of `lm()` ).

You can get standard diagnostic plots for most linear models with the `plot()` function. 
```{r}
plot(lm1) # diagnostic plots
```

You will see that observation 11 seems to have a lot of influence (Cook's distance > 1, high residual deviance). It might be worth seeing if our results hold up if we remove this observation. We can use our subsetting wizardry to make this easy.
```{r}
lm2 <- lm(rMax ~ Light, data=growth[-11,]) # we do not want to 11th row
summary(lm2)
```
(Note: if we were using separate vectors and not a data frame we would have had to have typed `lm(rmax[-11] ~ Light[-11])`, which is OK, but you can see that the possibilities for mistakes increase... )

Yup, the slope is reduced and is no longer significant when this observation is removed. So if this were real data, you would need to think about whether it was real or in error. But let us end by 

### Homework: try it on your own
Here are some data from Bolker's book showing the functional response of predators to Reed frog tadpole density. We are interested in knowing how predation rates increase with tadpole density. In this dataset, "Initial" is the initial density and "Killed" is the number of tadpoles killed by the predator.
```{r}
func <- data.frame(Initial=c(5, 5, 10, 10, 15, 15, 20, 20, 30, 30, 50, 50, 75, 75, 100, 100), 
				Killed=c(1, 2, 5, 6, 10, 9, 7, 10, 11, 15, 5, 21, 32, 18, 25, 35))
```

Using these data:

1.  plot the number killed against the initial density.
2.  add a column for the predation rate (=killed/initial density)
3.  find the average predation rates across all rows (=replicate aquaria)
4.  plot the predation rate  against the initial density.
5.  determine whether the predation rate statistically changes (linearly) with density

```{r, echo=FALSE, eval=FALSE}
#1
with(func, plot(x=Initial, y=Killed))
#2
func$PredRate <- with(func, Killed/Initial)
#3
summary(func$PredRate)
#4
with(func, plot(x=Initial, y=PredRate) )
#5
lm.pred <- lm(PredRate ~ Initial, data=func)
summary(lm.pred)
```

