---
title: 'Lab: Playing with deterministic functions'
author: "Jesse Brunner"
date: '`r format(Sys.Date())`'
output:
  pdf_document:
    toc: yes
  html_document:
    toc: yes
---


```{r, echo=FALSE}
knitr::opts_chunk$set(fig.width = 4, fig.height = 3)
```


Our goals in this lab are to: 

1. get a feel for a suite of important deterministic functions, seeing how their parameters change their shapes and behavior
2. create reference figures that you can refer to when using these functions

There is no short cut for doing the math
----------------------------------------

Most of the focus of this lab is plotting deterministic functions in R because 1) this is useful and 2) the labs are largely based on learning R. However, plotting functions is not a short cut for understanding them. In general you may be better off with a pencil and paper, just trying to figure out how an equation works. For instance, can you separate the equation into parts? What if you set one part or the other to zero? Are there parameter values where this happens? 

You would also do well to find inflection points by taking the derivative and setting it equal to zero. These points are useful. For instance, you might use them to find approximate values for your deterministic value when fitting a model to data. You should also have a sense of whether you are getting reasonable results from a fitting algorithm. R can calculate derivatives for you:
```{r D}
D( expression((x^3)/2), "x")
```
but as you will see it doesn't simplify and it doesn't always make it easy to see what's going on. There are better/clearer tools online (e.g., Wolfram Alpha and http://www.derivative-calculator.net/, which shows you the steps of its derivation). So try first on paper and lean on these tools to check your work or get through tricky bits. Once you really understand how a function works it will be a tool for life!


Plotting curves with `stat_function()`
------------------------------------

One of the best ways to understand how a deterministic function behaves and what the parameters do/mean is to plot it with different values over different ranges. The `stat_function()` function in the ggplot2 package is very helpful for this. 
We can plot any function we like. We might be interested in understanding the negative exponential function, for instance:
$$f(x) = a \times \exp(-bx)$$
where $a = 5$ and $b = 2$. We just give `ggplot` a data frame with a low and high value for the x-axis (`stat_function` automatically divides this span up into `n=101` chunks when plotting) and then apply `stat_function`, telling it what function to use to generate the y-values.
```{r Stat_Fun1}
library(ggplot2)
ggplot(data=data.frame(x=c(0,10)), aes(x=x) ) + 
			 	stat_function(fun=function(x)  5*exp(-2*x) )
```

We could also give `stat_function` the name of the function we wanted to plot, such as our `NegExp()` function from last time. 
```{r Stat_Fun2}
NegExp <- function(x, a = 1, b = 1) {
	stopifnot(x>=0, a>=0, b>=0)
	y <- a*exp(-b*x)
	return(y)
}

p <- ggplot(data=data.frame(x=c(0,10)), aes(x=x) )
p + stat_function(fun=NegExp)
```

Notice that we are seeing the default values of `NegExp()`, which are $a = 1, b = 1$. How do we change these? With the `args` argument.
```{r Stat_Fun3}
p + stat_function(fun=NegExp, args=list(b=3))
```

If we want to plot more than one curve on the same axes we need to apply multiple `stat_function`s to the plot, such as:
```{r guide_NegExp1}
p + stat_function(fun=NegExp, args=list(a=1, b=0.5)) +  
	stat_function(fun=NegExp, args=list(a=1, b=1)) +   
	stat_function(fun=NegExp, args=list(a=1, b=2)) +  
	stat_function(fun=NegExp, args=list(a=1/2, b=1)) + 
	labs(title = "Negative Exponential") + theme_bw()
```
To make these lines different colors and have a labeled legend we need to assign a label to the color aesthetic using the `aes()` function within each `stat_function()` call. This works basically as if we had a factor variable for the parameters and were assigning it to the color aesthetic.
```{r guide_NegExp2}
p + stat_function(fun=NegExp, args=list(a=1, b=0.5), aes(color="a=1, b=0.5")) +  
	stat_function(fun=NegExp, args=list(a=1, b=1), aes(color="a=1, b=1")) +   
	stat_function(fun=NegExp, args=list(a=1, b=2), aes(color="a=1, b=2")) +  
	stat_function(fun=NegExp, args=list(a=1/2, b=1), aes(color="a=1/2, b=1")) + 
	labs(title = "Negative Exponential") + theme_bw() + 
	scale_color_discrete("Parameters") # optional, but gives label to legend
```

(Bonus: try log-transforming the y-axis with `scale_y_log10()`. Does this match your expectation? Try doing the math (take the log of both sides of the equation) to ensure you remember why two lines are parallel.)

This is approach of adding lots of `stat_function()`s is certainly workable, but it does get a bit tedious and can lead to errors. The alternative is to just create your data frame with columns of x and the $a$ and $b$ parameters and then calculate what y is. The `expand.grid()` is very helpful for such things.
```{r guide_NegExp_expand.grid}
# create a data frame with all combinations of x, a, and b
df <- expand.grid(x=seq(0,10, length=101), 
									a=c(0.5,1), 
									b=c(0.5,1,2))
# add in the predictions
df$y <- with(df, NegExp(x=x, a=a, b=b))
head(df)
tail(df)
# add in a column with the levels of the parameters
df$Parameters <- with(df, paste("a=", a, ", b=", b, sep=""))
```
You can then plot this as if it were data.
```{r guide_NegExp3}
ggplot(data=df, aes(x=x, y=y, color=Parameters)) +
	geom_line() + 
	labs(title = "Negative Exponential") + theme_bw()
# Or alternatively
ggplot(data=df, aes(x=x, y=y, linetype=factor(a), color=factor(b),  group=Parameters)) +
	geom_line() + 
	scale_linetype("a = ") + 
	scale_color_discrete("b = ") +
	labs(title = "Negative Exponential") + theme_bw()
```

I think that this sort of figure is a handy reference. You may want to turn this into a pdf and keep it handy or print it out. It's relatively easy to do:
```{r NegExp_pdf, eval=FALSE}
ggsave(file="NegativeExponential.pdf", width = 8, height = 6) # width in inches, by default
```

More functions to explore
-------------------------

You should try to at least understand the behavior of, if not create a similar reference pdf of the following deterministic functions. See where they saturate, where the half-maximum shows up, where they wiggle, etc. The approach is the same. Try to code them yourself. If you have trouble with the math, see below.

**Hyperbolic**  
$$
\begin{aligned}
y & = \frac{a}{b+x} 
\end{aligned}
$$

**Michaelis-Menton**
$$
\begin{aligned}
y & = \frac{ax}{b+x} 
\end{aligned}
$$

**Holling Type III**
$$
\begin{aligned}
y & = \frac{ax^2}{b^2+x^2} 
\end{aligned}
$$

**Holling Type IV**
$$
\begin{aligned}
y & = \frac{ax^2}{b+cx+x^2} 
\end{aligned}
$$

**Monomolecular**
$$
\begin{aligned}
y & = a(1-exp(-bx)) 
\end{aligned}
$$

**Ricker**
$$
\begin{aligned}
y & = ax \times exp(-bx)
\end{aligned}
$$

**Logistic**
$$
\begin{aligned}
y & = \frac{exp(a+bx)}{1+exp(a+bx)}
\end{aligned}
$$


These are for your own benefit, so do them, but there is no need to send me anything unless you have questions or produce something really cool.




ifelse() for Piece-wise Functions 
--------------------------------

The `ifelse()` command in R is useful for constructing piece-wise functions.  Its basic syntax is `ifelse(condition, value_if_true, value_if_false)`, where `condition` is a logical vector (e.g., `x > 0`), `value_if_true` is a vector of alternatives to use if condition is `TRUE`, and `value_if_false` is a vector of alternatives to use if condition is `FALSE`.  If you specify just one value, it will be expanded (recycled in R jargon) to be the right length.  A simple example:
```{r ifelse_example}
x <- c(-25, -16, -9, -4, -1, 0, 1, 4, 9, 16, 25)
ifelse(x < 0, 0, sqrt(x)) # gives warnings because it evaluates sqrt(x)
# better alternative?
sqrt(ifelse(x<0, 0, x))
```

Here are some examples of using `ifelse()` to generate 

1. a simple threshold   
2. a Holling type I or “hockey stick”   
3. a more complicated piece-wise model that grows exponentially and then decreases linearly   
4. a double-threshold model   

```{r ifelse_plots}

ggplot(data.frame(x=c(0,5)), aes(x)) + 
stat_function(fun=function(x) ifelse(x < 2, 1/2, 2),
							aes(color="ifelse(x < 2, 1/2, 2)")) +   
stat_function(fun=function(x) ifelse(x < 2, 2 * x, 4), 
							aes(color="ifelse(x < 2, 2 * x, 4)")) +   
stat_function(fun=function(x) ifelse(x < 2, exp(x), 7*exp(-(x-2)/4)), 
							aes(color="ifelse(x < 2, exp(x), 7*exp(-(x-2)/4))")) +   
stat_function(fun=function(x) ifelse(x < 2, 1, ifelse(x < 4, 3, 5)), 
							aes(color="ifelse(x < 2, 1, ifelse(x < 4, 3, 5))")) +   
scale_colour_discrete("Piecewise model") + 
theme_bw() +   
theme(legend.position=c(0,1), 
			legend.justification=c(0,1), 
			legend.background=element_blank())

# Notice that last call to theme(). 
# I'm specifying the position and justification to move
# the legend position to the upper left of the graph 
# so that it doesn't squish the graph itself. 
# The part about the background is to prevent the  
# legend's white background from covering the lines.
# element_blank() is way to "delete" an element
```

The double-threshold example (nested `ifelse()` commands) probably needs more explanation.   In words, this command would read “if x is less than 2, set y to 1; otherwise (where x $\ge$ 2), if x is less than 4 (i.e., 2 $\leq$ x $<$ 4), set y to 3; otherwise (where x $\ge$ 4), set y to 5”.        


Homework: Which curve?
-------------------------
	
1. On page 96 in EMD, Bolker states that the logistic is popular because it is a simple sigmoid function, but then indicates that its "rational analogue," the Holling Type III functional response is also a simple sigmoid function. Does that mean they are interchangeable? Can you get a Holling type III curve to look like a logistic or vice versa? (Say, from 0 to 10.) If so, under what parameters? What about the hyperbolic and the negative exponential?

Send me one figure for each pair showing me how close you were able to get over the domain {0,10}. Make sure you label your axes, etc., and indicate the parameter combinations you used. Closest fit in the class gets a beer.

2. Describe and plot the deterministic function you think would relate your predictor (x-variable) to your response (y-variable). (This likely relates to the data you worked with and plotted last week.) Include a 1-2 sentence caption as you would in a manuscript.



```{r morefunctions, echo=FALSE, eval=FALSE}
Hyperbolic <- function(x, a = 2, b = 3) { a/(b+x) }  
MichMent <- function(x, a = 2, b = 1) { a*x / (b + x) }  
Holling3 <- function(x, a = 2, b = 1) { (a*x^2) / (b^2 + x^2) }  
Holling4 <- function(x, a = 2, b = 3, c = -1) { (a*x^2) / (b + c*x + x^2) }   
Monomolecular <- function(x, a = 2, b = 3) { a*(1 - exp(-b*x) ) }  
Ricker <- function(x, a = 2, b = 3) { a*x * exp(-b*x) }  
Logistic <- function(x, a = 0, b = 1) { exp(a+b*x) / (1 + exp(a+b*x) ) }  
```


```{r guide_Hyperbolic, echo=FALSE, eval=FALSE}

#### NEED TO ADJUST PARAMETES & COLORS
Hyp <- ggplot(data.frame(x=c(0,10)), aes(x)) +   
	stat_function(fun=Hyperbolic,  args = list(a=1, b = 0.5), aes(colour="a=1, b=0.5")) +   
	stat_function(fun=Hyperbolic,  args = list(a=1, b = 1), aes(colour="a=1, b=1")) +   
	stat_function(fun=Hyperbolic,  args = list(a=1, b = 5), aes(colour="a=1, b=5")) +   
	stat_function(fun=Hyperbolic,  args = list(a=1/2, b = 1), aes(colour="a=0.5, b=1")) +   
	stat_function(fun=Hyperbolic,  args = list(a=5, b = 1), aes(colour="a=2, b=1")) +
	scale_colour_discrete("Parameters") +   
	labs(title = "Hyperbolic") + theme_bw()
Hyp
```

```{r guide_MichMent, echo=FALSE, eval=FALSE}
MM <- ggplot(data.frame(x=c(0,10)), aes(x)) +   
	stat_function(fun=MichMent,  args = list(a=1, b=0.5), aes(colour="a=1, b=0.5")) +   
	stat_function(fun=MichMent,  args = list(a=1, b=1), aes(colour="a=1, b=1")) +   
	stat_function(fun=MichMent,  args = list(a=1, b=5), aes(colour="a=1, b=5")) +   
	stat_function(fun=MichMent,  args = list(a=1/2, b=1), aes(colour="a=0.5, b=1")) +   	stat_function(fun=MichMent,  args = list(a=2, b=1), aes(colour="a=2, b=1")) +
	scale_colour_discrete("Parameters") +   
	labs(title = "Michaelis-Menton") + theme_bw()
MM
```


```{r guide_Holling3, echo=FALSE, eval=FALSE}
H3 <- ggplot(data.frame(x=c(0,10)), aes(x)) +   
	stat_function(fun=Holling3, args = list(a=1, b=0.5), aes(colour="a=1, b=0.5")) +   
	stat_function(fun=Holling3, args = list(a=1, b=1), aes(colour="a=1, b=1")) +   
	stat_function(fun=Holling3, args = list(a=1, b=5), aes(colour="a=1, b=5")) +   
	stat_function(fun=Holling3, args = list(a=1/2, b=1), aes(colour="a=0.5, b=1")) +   
	stat_function(fun=Holling3, args = list(a=2, b=1), aes(colour="a=2, b=1")) +
	scale_colour_discrete("Parameters") +   
	labs(title = "Holling Type III") + theme_bw()
H3
```

```{r guide_Holling4, echo=FALSE, eval=FALSE}

### add in parameter c
H4 <- ggplot(data.frame(x=c(0,10)), aes(x)) +   
	stat_function(fun=Holling4,  args = list(a=1, b = 0.5), aes(colour="a=1, b=0.5")) +   
	stat_function(fun=Holling4,  args = list(a=1, b = 1), aes(colour="a=1, b=1")) +   
	stat_function(fun=Holling4,  args = list(a=1, b = 5), aes(colour="a=1, b=5")) +   
	stat_function(fun=Holling4,  args = list(a=1/2, b = 1), aes(colour="a=0.5, b=1")) +   	stat_function(fun=Holling4,  args = list(a=5, b = 1), aes(colour="a=2, b=1")) +
	scale_colour_manual("Parameters", values=c("orange", "green","black","darkgreen", "red"), breaks=c("a=1, b=0.5", "a=1, b=1", "a=1, b=5", "a=0.5, b=1", "a=5, b=1")) +   
	labs(title = "Holling Type IV") + theme_bw()
H4
```

```{r guide_Monomolecular, echo=FALSE, eval=FALSE}
Mono <- ggplot(data.frame(x=c(0,10)), aes(x)) +   
	stat_function(fun=Monomolecular,  args = list(a=1, b = 0.5), aes(colour="a=1, b=0.5")) +   
	stat_function(fun=Monomolecular,  args = list(a=1, b = 1), aes(colour="a=1, b=1")) +   
	stat_function(fun=Monomolecular,  args = list(a=1, b = 5), aes(colour="a=1, b=5")) +   
	stat_function(fun=Monomolecular,  args = list(a=1/2, b = 1), aes(colour="a=0.5, b=1")) +   	stat_function(fun=Monomolecular,  args = list(a=5, b = 1), aes(colour="a=2, b=1")) +
	scale_colour_manual("Parameters", values=c("orange", "green","black","darkgreen", "red"), breaks=c("a=1, b=0.5", "a=1, b=1", "a=1, b=5", "a=0.5, b=1", "a=5, b=1")) +   
	labs(title = "Monomolecular") + theme_bw()
Mono
```


```{r guide_Ricker, echo=FALSE, eval=FALSE}
Rick <- ggplot(data.frame(x=c(0,10)), aes(x)) +   
	stat_function(fun=Ricker,  args = list(a=1, b = 0.5), aes(colour="a=1, b=0.5")) +   
	stat_function(fun=Ricker,  args = list(a=1, b = 1), aes(colour="a=1, b=1")) +   
	stat_function(fun=Ricker,  args = list(a=1, b = 5), aes(colour="a=1, b=5")) +   
	stat_function(fun=Ricker,  args = list(a=1/2, b = 1), aes(colour="a=0.5, b=1")) +   	stat_function(fun=Ricker,  args = list(a=5, b = 1), aes(colour="a=2, b=1")) +
	scale_colour_manual("Parameters", values=c("orange", "green","black","darkgreen", "red"), breaks=c("a=1, b=0.5", "a=1, b=1", "a=1, b=5", "a=0.5, b=1", "a=5, b=1")) +   
	labs(title = "Ricker") + theme_bw()
Rick
```

```{r guide_Logistic, echo=FALSE, eval=FALSE}
Logist <- ggplot(data.frame(x=c(0,10)), aes(x)) +   
	stat_function(fun=Logistic,  args = list(a=1, b = 0.5), aes(colour="a=1, b=0.5")) +   
	stat_function(fun=Logistic,  args = list(a=1, b = 1), aes(colour="a=1, b=1")) +   
	stat_function(fun=Logistic,  args = list(a=1, b = 5), aes(colour="a=1, b=5")) +   
	stat_function(fun=Logistic,  args = list(a=1/2, b = 1), aes(colour="a=0.5, b=1")) +   	stat_function(fun=Logistic,  args = list(a=5, b = 1), aes(colour="a=2, b=1")) +
	scale_colour_manual("Parameters", values=c("orange", "green","black","darkgreen", "red"), breaks=c("a=1, b=0.5", "a=1, b=1", "a=1, b=5", "a=0.5, b=1", "a=5, b=1")) +   
	labs(title = "Logistic") + theme_bw()
Logist
```
