---
title: 'Stochastic distributions: EMD chapter 4'
author: Jesse Brunner (with contributions from Jacqui Frair & John Stella, and of
  course Ben Bolker
date: "February 10, 2015"
output: html_document
---


Our goals in this lab are to: 

1. get a feel for a suite of important _stochastic_ distributions, seeing how their parameters changer their shapes and behavior
2. understand the processes that might produce some of these distributions
3. create reference figures that you can refer to when using these distributions

Distribution Functions
-----------------------
R knows about lots of probability distributions. For each, it can generate random numbers drawn from the distribution ('deviates'); compute the cumulative distribution function and the probability distribution function; and compute the quantile function, which gives the x value such that $\int_0^x \! P(x) \, \mathrm{d} x$ (area under the curve from $0$ to $x$) is a specified value, such as 0.95. (Think about "tail areas" from standard statistics.)

The figure below shows for an arbitrary distribution the R commands showing the density function (`ddist()`), cumulative distribution function (`pdist()`), quantile function (`qdist()`), and random deviate function (`rdist()`) where "`dist`" is the name of the distributions (e.g., `binom`, `norm`, `t`,  `pois`, `chi`, etc.). 

```{r dgamma_dist_example, fig.width = 15, fig.height=5, echo=FALSE}
sh = 4 # shape paramter of gamma

op=par(mfrow=c(1,2),mar=c(5.1,9.1,0,0.5))
# set up left figure axes & scale
curve(dgamma(x,shape=sh),from=0,to=20,ylab="",lwd=2,axes=FALSE)
axis(side=1,labels=FALSE)
axis(side=2,labels=FALSE)
box()

# plot a polygon of area under the curve
xvec = seq(0,5,length=100)
polygon(c(xvec,rev(xvec)),c(rep(0,length(xvec)),dgamma(rev(xvec),shape=sh)),col="gray",border=NA)
# plot the dgamma curve
curve(dgamma(x,shape=sh),from=0,to=20,lwd=2,add=TRUE)
# add a vertical line at x=5
abline(v=5,lty=3)
# add horizontal line showing dgamma(x=5)
abline(h=dgamma(5,shape=sh),lty=2,lwd=2)

# add marginal text
mtext(side=1,line=1,at=5,expression(x[0]),cex=1.5)
mtext(side=2,at=dgamma(5,shape=sh),las=1,expression(ddist(x[0])),
      line=1.5,cex=1.5)
mtext(side=2,at=0.0,adj=0,"Probability density",cex=1.5,line=1.5)
# add label for pdist line
text(5,0.2,expression(pdist(x[0])),cex=1.5,pos=4,col="darkgray")

# add random gamma draws
set.seed(1001)
points(rgamma(10,shape=sh),rep(0,10),cex=1.5)
# add text and arrow pointing to random draws
text(11.7,0.03,"rdist(10)",adj=0,cex=1.5)
arrows(10.8,0.023,6.6,0.008,lwd=2)

# on right plot, setup axes and scales, plot pgamma() line
curve(pgamma(x,shape=sh),from=0,to=20,ylab="",lwd=2,axes=FALSE)
axis(side=1,labels=FALSE)
axis(side=2,labels=FALSE)
box()

# add horizontal line at pgamma(q=5)
abline(h=pgamma(5,shape=sh),lty=2,lwd=2)
# add vertical line for qgamma(p=0.95)
abline(v=qgamma(0.95,shape=sh),lty=4,lwd=2)
# add vertical line where x=5
abline(v=5,lty=3)
# add line segment where pdist = qgamma(p=0.95) line
segments(par("usr")[1],0.95,qgamma(0.95,shape=sh),0.95,lty=4,lwd=2)

# add marginal text
mtext(side=2,at=pgamma(5,shape=sh),las=1,expression(pdist(x[0])),
      line=1.5,cex=1.5)
mtext(side=2,at=0.95,las=1,0.95,
      line=par("mgp")[2],cex=1.5)
mtext(side=1,line=1.5,at=5,expression(x[0]),cex=1.5)
mtext(side=1,at=qgamma(0.95,shape=sh),text="qdist(0.95)",line=1.5,
      cex=1.5,adj=0.1)
mtext(side=2,at=-0.05,adj=0,"Cumulative distribution",cex=1.5,line=1.5)

```


To illustrate how we can use these functions, let's consider the Binomial distribution, with which you are probably familiar. It is essentially a coin-flip where one outcome, say heads, is "success" or an "event."

`rbinom(n, size, p)` gives `n` random samples each with `size` binomial trials (= coin flips) per sample, and `p` probability of success in each trial. For example, to simulate an experiment where 10 people (10 samples) each flip a coin 8 times (= 8 trials) with a fair coin (probability of success [1] is 0.5) we would use the following command:
```{r binom_coin}
rbinom(n = 10, size = 8, p = 0.5)
```

A simulation of 200 values from a binomial distribution with `size` = 12 (12 binary trials) and `p` = 0.15 per trial would look like this:

```{r, warning=FALSE, message=FALSE, fig.width=5, fig.height=4}
successes <- rbinom(n=200, size=12, p=0.15)

library(ggplot2)
qplot(successes, xlim = c(0,15))
```

`dbinom(x,size,p)` gives the value of the probability distribution function (pdf) for any given value of $x$.  Since the binomial is discrete, $x$ has to be an integer, and the pdf is just the probability of getting that many successes; if you try dbinom with a non-integer x, you’ll get a zero and a warning. (Note: For a continuous distribution the analog is the probability _density_ function, which calculates the probability density within some range $[a, b]$).

```{r}
# This is the probability of exactly 2 successes out of 15 trials, 
# with a 15% chance of sucess on each trial 
dbinom(x = 2, size = 15, prob = 0.15) 
```

A plot illustrates the general pattern of what the density function does.
```{r dbinom, fig.width=5, fig.height=4}
x <- 0:15
qplot(x, y = dbinom(x, size = 15, prob = 0.15), 
			xlab = "number of successes", ylab = "probability") # default geom = "point"
# Or alternatively
qplot(x, ymax=dbinom(x, size = 15, prob = 0.15), ymin = 0, 
			geom="linerange", xlab = "number of successes", ylab = "probability")
```

These plots show the probability of getting $x$ successes given 15 trials and a probability of 0.15. Notice that they sum to 1.


`pbinom(q, size, p)` gives the value of the _cumulative_ distribution function (cdf) at `q`. That is, it gives the probability of getting `q` or fewer successes. (This should sound sort of familiar. It is close to the definition of the p-value, only in that case we are interested in the probability of `q` _or more_ successes, which is generally `1 - pbinom()`.) For example:

```{r}
# The probability of get 2 or fewer successes (i.e., 0, 1, or 2)
pbinom(q = 2, size = 15, prob = 0.15) 
# We can get the same result by summing up the probabilities of getting 0, 1, and 2 successes
sum(dbinom(x=0:2, size = 15, prob = 0.15)) 
```

A plot can illustrate what cumulative probability distribution looks like.
```{r pbinom, fig.width=5, fig.height=4}
qplot(x, y = pbinom(q = x, size = 15, prob = 0.15), ylim = c(0,1), 
			xlab = "number of successes", ylab = "cummulative probability")
# Or alternatively, with vertical lines, like a histogram
qplot(x, ymax = pbinom(q = x, size = 15, prob = 0.15), ymin = 0, 
			geom="linerange", ylim = c(0,1), 
			xlab = "number of successes", ylab = "cummulative probability")
```

These plots then show the cumulative probability of getting $x$ or fewer successes. See that they quickly converge on 1.

Finally, `qbinom(p, size, prob)` gives the quantile function, where `p` is the proportion of the area under the pdf, or identically, the value of the cdf you want to evaluate. The quantile function returns the value of $x$ associated with that cumulative probability, i.e., the value of $x$ such that $P(x \le p) = q$. (Note that the quantile function $Q$ is the inverse of the cumulative distribution function $C$: if $Q(p) = q$ then $C(q) = p$.)

```{r qbinom, fig.width=5, fig.height=4}
# There is a 95% probability that in 15 trials, 
# each with 0.15 chance of success, 
# there will be this many or fewer successes
qbinom(p = 0.95, size = 15, prob = 0.15) 

qplot(x=c(0,1), stat="function", fun = qbinom, geom = "step",
			args = list(size = 15, prob = 0.15), 
			xlab = "quantile (or proportion of area under pdf)", ylab = "number of successes")
```

Notice that because the binomial deals with discrete outcomes, there are a range of cumulative probabilities associated with each number outcomes (i.e., it is a jagged curve).


These four functions exist for each of the distributions R has built in: e.g., for the normal distribution they are `rnorm()`, `pnorm()`, `dnorm()`,and `qnorm()`. Each distribution has its own set of parameters as appropriate. For example, `pnorm(x, mean, sd)`. Use `?rnorm` for more information.

---------------

**Test yourself**  
Use the above commands to work out the following for a binomial distribution with 10 trials and a success probability of 0.2.

* Pick 8 random values and sort them into increasing order. (If you `set.seed(1001)` `r set.seed(1001)` just before your calculations you should get `r sort( rbinom(8, 10, 0.2) )`).

* Calculate the probabilities of getting 3, 4, or 5 successes. (Answer: `r dbinom(3, 10, 0.2)`, `r dbinom(4, 10, 0.2)`, and `r dbinom(5, 10, 0.2)`, respectively)

* Calculate the probability of getting 5 or more successes. (Answer: `r 1-pbinom(4, 10, 0.2)`)

* What tail values would you use to test against the (two-sided) null hypothesis that p = 0.2? (Use `qbinom()` to get the answer, and use `pbinom(0:10, size = 10, prob = 0.2)` and `pbinom(0:10, size = 10, prob = 0.2, lower.tail=FALSE)` to check that your answer makes sense.)

--------------

Plotting discrete stochastic distributions
------------------------------------------

Once again, it is useful to create a visual reference for these functions, particularly the pdfs (`dbinom()`), showing how parameter values alter the shape and look of the distributions. We can use the `stat_function` approach we used in the previous lab to plot discrete distributions. We need only specify "`n`," the number of points to evaluate the function along. (Think about why that is.)

```{r dbinom_stat_function, fig.width=5, fig.height=4}
qplot(x=0:15, n = length(x), stat = "function", fun = dbinom, 
			args = list(size=15, prob = 0.5), 
			xlab = "successes", ylab = "probability", main = "probability = 0.5")
```

Or we can generate the data in a data frame, and then plot it. We will take this approach because a) it is  worth learning how to do so and b) it make creating a faceted or overlayed plot much simpler. So let's see how to generate the probabilities of success (pdfs) for probability of success = 0.1, 0.3, and 0.5 with 5, 15, and 50 trials. We will make use of the `expand.grid()` function, which is designed for this very purpose. It creates a data frame with all combinations of each vector we give it.

```{r create_df_5}
df.5 <- expand.grid(x = c(0:5), trials = 5, prob = c(0.1, 0.3, 0.5))
df.5 # see how it is structured? 
```

See how this works? If not, try futzing with the function until you do. Or ask your neighbor questions.

OK. So we've created one data frame for five trials. Now let's create them for 15 and 50 trials and then combine them all together.

```{r create_df}
df.15 <- expand.grid(x = c(0:15), trials = 15, prob = c(0.1, 0.3, 0.5))
df.50 <- expand.grid(x = c(0:50), trials = 50, prob = c(0.1, 0.3, 0.5))
# Now, to combine them all
df <- rbind(df.5, df.15, df.50)
```

Of course we could have done this in a single command:
```{r create_df_alt}
df <- rbind( 
	expand.grid(x = c(0:5), trials = 5, prob = c(0.1, 0.3, 0.5)),   
  expand.grid(x = c(0:15), trials = 15, prob = c(0.1, 0.3, 0.5)),   
  expand.grid(x = c(0:50), trials = 50, prob = c(0.1, 0.3, 0.5))   
	)
```

Anyway, we now have a data frame, `df`, with every combination of `x` and `prob`, along with labels for the number of `trials`. We still need to calculate the y-values, the probability of each outcome (`x` success) given the probability of success (`prob`), but this is pretty straightforward. We simply need to supply the right values of size and prob t the `dbinom()` function:
```{r create_df_y}
df$y <- with(df, dbinom(x, size = trials, prob = prob) )
```

It is then simple to plot this with different facets for each combination of parameters (9 facets, right?).

```{r plot_dbinom_facets, fig.height=4}
ggplot(data = df, aes(x = x, y = y, ymax = y, ymin = 0)) + 
	geom_point() + geom_linerange() + 
	facet_grid(facets = prob ~ trials ) + 
	labs(x="Number of successes", y = "Probability ")
```

But wait! We have the same x-limits on each column, even though the maximum number of successes varies. We can let the x-axis be "free" to allow different limits and create a prettier, more honest figure.

```{r plot_dbinom_facets_xfree, fig.height=4}

ggplot(data = df, aes(x = x, y = y, ymax = y, ymin = 0)) + 
	geom_point() + geom_linerange() + 
	facet_grid(facets = prob ~ trials, scales="free_x") + 
	labs(x="Number of successes", y = "Probability ")
```

We could also allow the y-axis to be free with `scales="free_xy"` or `scales="free_y"`.

Lastly, saving figures in your script can be helpful and handy. Try the `ggsave()` function, which by default saves the last _displayed_ ggplot figure to whatever file format matches the file extension.
```{r ggsave, eval=FALSE}
ggsave(file="binomial_probs.pdf", width=6.5, height=5) # units are inches by default
```



Homework: More functions to explore
------------------------------------

Using the same basic tools, you should play with and create similar visual guides for the several other stochastic distributions. These include the other discrete distributions:

Distribution  | R function
------------- | -------------
Poisson  | `dpois(x, lambda)`
negative binomial  | `dnbinom(x, size, prob, mu)`

And the continuous distributions:

Distribution  | R function
------------- | -------------
normal  | `dnorm(x, mean, sd)`
log normal  | `dlnorm(x, meanlog, sdlog)`
Gamma  | `dgamma(x, shape, rate, scale)` # scale = 1/rate
beta  | `dbeta(x, shape1, shape2)`







