---
title: "Tools of the Ecological Detective"
author: "Jesse Brunner"
date: "1/10/2017"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Goals of the class

- Learn to fit models to data
     +  meaningful, mechanistic or phenomenological 
     +  representing hypotheses of interest
     +  _multiple_
- Evaluate evidence for/against models under different frameworks
     +  hypothesis testing w/ p-values
     +  Likelihood
     +  Bayesian (just barely)
- Present your findings
- Use R to conduct reproducable research



## Fitting models to data

Data = Signal + Noise
```{r, echo=FALSE}
library(ggplot2)
df <- data.frame(x=rnorm(30, mean=10, sd=5))
df$y <- 20 + 5*df$x + rnorm(30, mean=0, sd=15)
ggplot(df, aes(x=x,y=y)) + 
	geom_point() + 
	geom_abline(intercept = 20, slope=5, color="blue") + 
	geom_segment((aes(xend=x, yend=20 + 5*df$x)), color="red")
```

Focus on size of the Signal (relative to precision of estimate)

## My role in this class

You can expect me to:

- Provide resources, context, and interesting challenges 
- Be helpful, courteous, and (as much as possible) responsive to your needs 
    + Office hours for life 
    + Please give me 24h to get back to you on Piazza 
- Acknowledge the limits of my undertanding (I'm still learning, too)

What you cannot expct of me: 

- Answers, at least not always
    + I may not know/be certain (this is an evolving field)
    + It may not be good for your learning 
- Doing your work for you (code, finding right question/analysis)

## Your role in this class

I expect you to:

- Try and understand what you are doing/reading 
- Talk to your classmates about your understanding
- Help each other (=key to learning!)
- Ask me questions when you are uncertain (or just to be sure)
- Contribute useful resources you find
- Complete the assignments on time (=a lot of time outside of lecture/lab) 
- Keep your damn code clean and readable!
- Be prepared for your follow up activity / presentation on Special Topic

## Rules for providing feedback 

- Be curteous... we are all in this together 
- Take the time to do it right. Are three word reviews helpful? No! 
- Be timely
- See what you can learn from others

## Rules for student follow up / presentations

You should 

- have 1--3 lesson(s) you want your audience to _learn_ 
- be interactive and engaging with activities for students to _do_
- minimize the use of R or at least coding in front of the class  
- be well prepared 
     - read beyond the chapter! 
     - sort out your questions and highlight things that tripped you up 
     - try out your activity, etc., before hand (we are not your guinea pigs...unless it's fun)
